---
title: "State-by-State Crime Analysis Using Hierarchical Clustering"
author: "Sejal Kankriya"
date: '2023-02-20'
output:
  pdf_document: default
  html_document:
    df_print: paged
always_allow_html: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(cluster)
library(dplyr)
library(magrittr)
library(ggplot2)
library(plotly)
library(data.table)
library(caret)
library(ggbiplot)
library(tidyr)
library(cowplot)
library(ggdendro)
library(factoextra)
```

# USArrests Dataset and Hierarchical Clustering

Consider the "USArrests" data. It is a built-in dataset you may directly
get in RStudio. Performing hierarchical clustering on the observations (states)

```{r}
head(USArrests)
```

Using hierarchical clustering with complete linkage and
Euclidean distance, cluster the states

```{r}
set.seed(8)

data("USArrests")

# Load the data
data <- USArrests

# Omiting the missing values if any
data <- na.omit(data)

# Computing the distance matrix using euclidean method
distance_matrix <- dist(data, method = "euclidean")

# Performing hierarchical clustering
hc.complete <- hclust(distance_matrix, method = "complete")

# Plot the dendrogram
plot(hc.complete, main = "Complete Linkage",
    xlab = "", sub = "", cex = .5, hang = -1)
```

```{r}
# ggplot
ggdendrogram(hc.complete, segements=TRUE, labels=TRUE, leaf_labels = TRUE, rotate=FALSE, theme_dendro = TRUE) +
 labs(title='Complete Linkage')
```

Cut the dendrogram at a height that results in three distinct
clusters and interpreting the clusters

```{r}
# Determining the cut tree
clusters <- cutree(hc.complete, 3)
clusters
```

```{r}
table(clusters)
cat("\nStates belonging to Cluster 1\n")
subset(row.names(USArrests), clusters == 1)
cat("\nStates belonging to Cluster 2\n")
subset(row.names(USArrests), clusters == 2)
cat("\nStates belonging to Cluster 3\n")
subset(row.names(USArrests), clusters == 3)
```

Cluster 1 contains the states with higher levels of violent crimes and
arrests, such as California, New York, Florida, and Illinois. Cluster 2
includes states that have moderate levels of violent crimes and arrests,
such as Arkansas, Georgia, and Tennessee. Finally, Cluster 3 consists of
states with lower levels of violent crimes and arrests, such as Maine,
Montana, and Vermont.

Hierarchically cluster the states using complete linkage and
Euclidean distance, after scaling the variables to have standard
deviation one. Obtaining three clusters.

```{r}
set.seed(6)
data("USArrests")
# Load the data
data <- USArrests
# Omiting the missing values if any
data <- na.omit(data)
# Standardizing the data before clustering
data_scaled <- scale(data)
# Computing the distance matrix using euclidean method
distance_matrix <- dist(data_scaled, method = "euclidean")
# Performing hierarchical clustering
hc.complete_scaled <- hclust(distance_matrix, method = "complete")
# Plot the dendrogram
plot(hc.complete_scaled, main = "Complete Linkage",
    xlab = "", sub = "", cex = .5, hang = -1)
```

```{r}
# ggplot
ggdendrogram(hc.complete_scaled, segements=TRUE, labels=TRUE, leaf_labels = TRUE, rotate=FALSE, theme_dendro = TRUE) +
 labs(title='Complete Linkage')
```

```{r}
# Determining the cut tree
scaled_cutree <- cutree(hc.complete, 3)
scaled_cutree
```

```{r}
table(scaled_cutree)
cat("\nStates belonging to Cluster 1\n")
subset(row.names(USArrests), scaled_cutree == 1)
cat("\nStates belonging to Cluster 2\n")
subset(row.names(USArrests), scaled_cutree == 2)
cat("\nStates belonging to Cluster 3\n")
subset(row.names(USArrests), scaled_cutree == 3)
```

The 50 states in the USArrests dataset have been separated into three distinct clusters based on their similarities and differences, according to the output of the cutree() tool. 

Alabama, Alaska, Georgia, Louisiana, Mississippi, North Carolina, South Carolina, and Tennessee are all part of Cluster 1. When the clusters are interpreted, we can observe that Cluster 1 predominantly consists of states in the United States' Southeastern area, such as Alabama, Georgia, and Louisiana, which have relatively high crime rates across all four categories examined in the dataset (assault, murder, rape, and robbery). 

Arizona, California, Colorado, Florida, Illinois, Maryland, Michigan, Nevada, New Mexico, New York, and Texas are part of Cluster 2. 
These states have intermediate rates of violent crime, murder, rape and assault arrest rates.

Cluster 3 consists of the remaining 32 states, which had lower rates of violent crime and arrest rates for murder, rape, and assault when compared to the other states in the dataset.